# **Analysis of longitudinal data**

*This is the last chapter of the IODS course. I will do analysis on a longitudinal data set and interpret the results.*

```{r}
date()

library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)

BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
RATS <- read.csv(file = "C:/LocalData/inkaronk/IODS-project/RATS.csv", header = TRUE)


RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))

RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4))) 

```

## RATS data
```{r}
glimpse(RATSL)

colnames(RATSL)
```

RATS data is from a nutrition study that had three groups of rats who were put on different diets. It has information about each groups body weight from 9 weeks.
RATS data set that has been turned in to long format has 192 rows and 5 columns. Column names are "ID", "Group", "WD", "Weight" and "Time".

### Graphical displays of longitudinal data
Let's look at the graphical displays of weight growth over time for both non-standardized and standardized RATS data.

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

From the first (non-standardized) graph we can see that the weight of the rats increases over time in each group, but the grams are different: In the group 1 rats weigh less than in group 2 and 3. There are more variety in group 2, but it's quite similar with group 3.


```{r}
RATSL <- RATSL %>%
  group_by(Group, Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

glimpse(RATSL)

ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "Standardized weight")

```


This graph explains same analysis but with the standardized version of used variables. It shows more exact results, the weights of the rats are more divided in each group. This graph shows also that the weight of some rats decreased, the first one showed only increasing weight trends.


### Summary measure analysis of longitudinal data
```{r}
n <- 20

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

glimpse(RATSS) # 36:4, Group, Time, mean, se

ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=1) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.2)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

This is the summary that shows mean profiles for each time points and the standard error of mean. It's quite similar with the first graph, group 1 has lower values and groups 2 and 3 are more similar, when group 2 has bit more variety in it.

### Choosing Summary Measures and Applying the Summary Measure Approach
```{r}
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

glimpse(RATSL8S)

ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time")
```

First we need to find the outliers, and we can see that there are outliers in groups, biggest one in 2. In group 2 the outlier is bigger than average (around 590) and in group 1 and 3 it's below the average (around 240 in 1 and around 490 in 3). Let's delete the outliers.

```{r}
RATSL8S1 <- RATSL8S %>%
  filter((Group==1 & mean > 260)|(Group==2 & mean < 500)| (Group==3 & mean > 520))

ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time")
```

### Anova

```{r}

RATSL8S2 <- RATSL8S1 %>%
  mutate(baseline = RATS$WD)

fit <- lm(mean ~ Group, data = RATSL8S2)

anova(fit)
```

With anova we can look at the difference between groups. F-value is statistically significant, which means that there is difference between groups.

## BPRS data

```{r}
glimpse(BPRSL)

colnames(BPRSL)
```

BPRS data has information about 40 male participants who were divided into two groups. Each participant was rated on the brief psychiatric rating scale (BPRS, 1-7) and then they got treatment for 8 weeks.
BPRS data set that has been turned in to long format has 400 rows and 5 columns. Column names are "treatment", "subject", "weeks", "bprs" and "week".
Let's look at the plot of individual treatment profiles.

```{r}
dim(BPRSL)

ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "week", breaks = seq(0, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = subject)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "week", breaks = seq(0, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "bottom")
```

Here we can graphically see how the treatment affected bprs during 8 weeks. There were more variety in treatment 2.


### The Random Intercept Model and Random Slope Model with interaction

```{r}
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref)

BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref1)

anova(BPRS_ref1, BPRS_ref)

BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)

anova(BPRS_ref2, BPRS_ref1)

ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = subject)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "week", breaks = seq(1, 4, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

Fitted <- fitted(BPRS_ref2)

BPRSL <- BPRSL %>% mutate(Fitted)

ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
  geom_line(aes(linetype = subject)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "week", breaks = seq(1, 4, 8)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "bottom")

```

First I did random intercept model for week and treatment variables that showed how the linear regression fir for each subject. Then I did random intercept and random slope model that shows individual differences and effect of time. And last with week x treatment interaction. First graph shows observed values and second one fitted values. As you can see, the fitted values are more easier to read, because they are more clear.